Can Symmetric Cryptography Be Liberated from the Von Neumann Style?

### 2024-10-14 paper.tex
cryptographic primitive

As intimidating as cryptographers are, the essence of modern symmetric
cryptography primitives can in fact be distilled in a few paragraph. A
primitive can be understood as a purely functional programs. It takes
two inputs, the input ``cleartext'' and a list of ``round subkeys''
(which are derived from a single key through a non-performance
critical process called ``key schedule'' that we shall ignore here).
The output is the ``ciphertext''. The cleartext, the ciphertext and
each individual round subkeys are processed as ``blocks'' of binary
data. The block size is of order of a hundred bits (typically, 64, 128
or 256 bits).

@@

Moreover, ciphers are now, by and large, designed to be run
efficiently on software. The era of hardware-based cryptography is
long gone, even for embedded
platforms~\citep{buchanan:nist-lightweight-crypto}. To achieve
high-throughput, ciphers are typically designed so as to admit a
bitsliced implementation. Bitslicing enables a form of parallel
execution \emph{within} machine words. We thus gain access to a form
of CPU-level ``scale out'' to larger machine words, in particular
through single-instruction multiple-data (SIMD) instructions. Doubling
the size of machine words yields nearly twice the throughput, without
any effort. Evidences tend to suggest that cryptographers obtain
bitsliced designs through sheer intellectual might. It is in fact the
sole purpose of the traditional ``Section 4'' scene to exhibit a
bitsliced witness in all the gory details.

@@

Bitslicing is the cryptographer's jargon for such a representation
change. In the case of \Skinny{}, we observe that only \coqe!shiftrows_!
imposes a strict composition order of \coqe!Rows.T! followed by
\coqe!Cols.T!. Aside from that, we can commute \coqe!Slice8.T! out the
composition. We would thus have the type \coqe!Rows.T (Cols.T bool)!
representing 16 bits and fitting snugly in a 32 bits machine word. The
overall composition \coqe!Slice8.T (Rows.T (Cols.T bool))! would take
8 registers in total, imposing a very reasonable amount of register
pressure to the CPU.
%
In effect, to return back to the jargon of compiler designers, we
propose to turn a structure-of-arrays (SoA) into an
array-of-structures (AoS), only we are dealing with bit-level
quantities here.

>>>>

As intimidating as cryptographers are, the essence of modern symmetric
cryptographic primitives can in fact be distilled in a few paragraph. A
primitive can be understood as a purely functional programs. It takes
two inputs, the input ``cleartext'' and a list of ``round subkeys''
(which are derived from a single key through a non-performance
critical process called ``key schedule'' that we shall ignore here).
The output is the ``ciphertext''. The cleartext, the ciphertext and
each individual round subkeys are processed as ``blocks'' of binary
data. The block size is of order of a hundred bits (typically, 64, 128
or 256 bits).

@@

Moreover, ciphers are now, by and large, designed to be run
efficiently on software. The era of hardware-based cryptography is
long gone, even for embedded
platforms~\citep{buchanan:nist-lightweight-crypto}. To achieve
high-throughput, ciphers are typically designed so as to admit a
bitsliced implementation. Bitslicing enables a form of parallel
execution \emph{within} machine words. We thus gain access to a form
of CPU-level ``scale out'' to larger machine words, in particular
through single-instruction multiple-data (SIMD) instructions. Doubling
the size of machine words yields nearly twice the throughput, without
any effort. Evidence tends to suggest that cryptographers obtain
bitsliced designs through sheer intellectual might. It is in fact the
sole purpose of the traditional ``Section 4'' scene to exhibit a
bitsliced witness in all the gory details.

@@

Bitslicing is the cryptographer's jargon for such a representation
change. In the case of \Skinny{}, we observe that only \coqe!shiftrows_!
imposes a strict composition order of \coqe!Rows.T! followed by
\coqe!Cols.T!. Aside from that, we can commute \coqe!Slice8.T! out of the
composition. We would thus have the type \coqe!Rows.T (Cols.T bool)!
representing 16 bits and fitting snugly in a 32 bits machine word. The
overall composition \coqe!Slice8.T (Rows.T (Cols.T bool))! would take
8 registers in total, imposing a very reasonable amount of register
pressure on the CPU.
%
In effect, to return back to the jargon of compiler designers, we
propose to turn a structure-of-arrays (SoA) into an
array-of-structures (AoS), only we are dealing with bit-level
quantities here.
